---
title: "Bootstrapping Exercise"
output: html_notebook
# render("bootstrap/RE_boot.Rmd")
---

<a href="../4-2-boot.html">Earlier</a> we looked at how we could apply bootstrapping to a linear model. We generated bootstrap standard errors for the OLS coefficient estimates. Here we will apply bootstrapping to the models to `ChickWeight` data. 

Let $weight_{ijt}$ be the weight of chick $i$ in Diet group $j$ observed in time $t$.  We will apply bootstrapping to the following models; 

- Model 1: linear time trend $$weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + \varepsilon_{ijt}$$ 

- Model 2: linear time trend with Chick fixed effects $$ weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + \alpha_i+ \varepsilon_{ijt}$$

- Model 3: linear time trend with Chick random effects $$ weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + v_{ijt}, \:\: v_{ijt} = \alpha_i+ \varepsilon_{ijt}$$


Load libraries and data.

```{r, warning=FALSE, message=FALSE}
    library(dplyr)
    library(ggplot2)
    library(broom)
    library(tidyr)
    library(mosaic)
    library(nlme)
    library(lme4)
    library(knitr)
load("movies2.RData")
```

### Define functions used for bootstrapping linear models 

```{r}
getFormula <- function(ols, lmer=FALSE) gsub("()","", ifelse(!lmer, ols$call[2], ols@call[2])) 

getDependentVar <- function(ols, lmer=FALSE) {
  str <- getFormula(ols, lmer=lmer) 
  gsub(" ","", substr(str, 1, (regexpr("~",str)[1]-1)))　 
}


run_ols_boot <- function(lm_rlt, num_do = 5000) {
  
  # calculate the standard deviation of the residuals
  N <- length(lm_rlt$residuals)
  sd_res <- (sum(lm_rlt$residuals^2)/lm_rlt$df.residual) %>% sqrt()
  dep_var <- getDependentVar(lm_rlt)

  do(num_do) * 
    ({  
        data_bt <- lm_rlt$model
        # replace the dependent variable with its bootstrap counterpart
        data_bt[[dep_var]] <- lm_rlt$fitted.values +    #  the predicted component
          + rnorm(N, mean = 0, sd = sd_res)     #  random draws from the error distribution 
         
        # run the OLS model with the same formula but with a new, bootstrap dataset  
        ols_bt <- lm(as.formula(getFormula(lm_rlt)), data = data_bt)  
        coef(ols_bt)  # get coefficients 
    }) 
}

```


```{r} 
# function generating a matrix of ones 
ones <- function(r,c) matrix(c(rep(1,r*c)),nrow=r,ncol=c)

# confidence interval by bootstrapping
# version 1 
ci_ver1 <- function(est_bt, alpha = 0.05) {
  # est_bt: bootstrap estimates with  row = boot replications, col = coefficients 
  apply(est_bt, 2, function(x) quantile(x, c(alpha/2, 1 - alpha/2))) %>% t() 
}

# version 2 
ci_ver2 <- function(est_sample, bt_sd, alpha = 0.05) {
  # est_semple: sample estimate vector 
  # bt_sd: bootstrap sd estimates of est_sample-vector  
  cbind('2.5%' = est_sample - 1.96 * bt_sd, 
      '97.5%' = est_sample + 1.96 * bt_sd) 
}
  
# bootstrap p-value 
bt_p_val <- function(est_sample, est_bt) {
  # est_semple: sample estimate vector 
  # est_bt: bootstrap estimates with  row = boot replications, col = coefficients 
  
  est_bt_long <- est_bt %>% data.frame() %>% gather()  
  
  est_bt_long <- est_bt_long %>%
    mutate(
        center_var = kronecker(ones(nrow(est_bt),1), matrix(est_sample, nrow=1)) %>% c(),
        extremes = abs(value - center_var) >= abs(center_var)
    )
    
  p_val <- est_bt_long %>% 
    group_by(key) %>% 
    summarise(p_val = sum(extremes)/nrow(est_bt))
  return(list(p_val=p_val, df_long=est_bt_long))
}

```

```{r}
# histogram visualization  
coeff_bt_histogram <- function(est_sample, est_bt, centering =FALSE) { 
  # est_semple: sample estimate vector 
  # est_bt: bootstrap estimates with  row = boot replications, col = coefficients 

  est_bt_long <- bt_p_val(est_sample, est_bt)$df_long
  
  if (centering) {
    est_bt_long <- est_bt_long %>%
      mutate(
          key = paste(key, " - center"),
          value = value - center_var,
          fill_var = extremes
      )
    legend_lab <- "Extremes: | value - center | > | center |"
    x_lab <- "value - center"
  } else {
    est_bt_long <- est_bt_long %>%
      mutate(
        sign = kronecker(ones(nrow(est_bt),1), matrix(ifelse(est_sample>0,1,-1), nrow=1)) %>% c(), 
        fill_var = value * sign <= 0
      )
    legend_lab <- "Crossing zero?"
    x_lab <- "value"
  }
  
  est_bt_long %>% 
    ggplot(aes(x = value, fill = fill_var)) +
    geom_histogram(color = "white", bins = 40) + theme(legend.position="bottom") +
    facet_wrap(~key, scales = "free") + labs(fill = legend_lab, x = x_lab)
}
```


```{r}
run_lmer_boot <- function(lmer_rlt, num_do = 5000) {
  # Randome effects (RE) model bootstrapping (random intercepts, not random slopes) 
  
  rlt <- list()
  rlt$model <- lmer_rlt@pp$X        # model data for the part of fixed coefficients
  N <- length(residuals(lmer_rlt))
  
  rlt$fitted_no_RE <- rlt$model %*% matrix(fixef(lmer_rlt), ncol=1)
  RE_vals <- lmer_rlt@flist[[1]] %>% unique()  #  RE variable values
  N_RE <- length(RE_vals)                 # number of RE variable values
  rlt$RE_idx <- rep(NA, N)
  for (i in 1:N_RE) rlt$RE_idx[which(lmer_rlt@flist[[1]] == RE_vals[i])] <- i  # index of RE variable values
 
  sd_res <-  sigma(lmer_rlt)        # standard deviation of the residuals
  sd_RE <-  lmer_rlt@theta * sd_res   # standard deviation of RE
  dep_var <- getDependentVar(lmer_rlt, lmer=TRUE) 

  do(num_do) * 
    ({  
        data_bt <- data.frame(lmer_rlt@frame)
    
        # replace the dependent variable with its bootstrap counterpart
        data_bt[[dep_var]] <- rlt$fitted_no_RE +          # the predicted component by fixed coefficients 
          + rnorm(N_RE, mean = 0, sd = sd_RE)[rlt$RE_idx] + # random draws of tge RE distribution
          + rnorm(N, mean = 0, sd = sd_res)               # random draws of the residual distribution
         
        # run the RE model with the same formula but with a new, bootstrap dataset  
        lmer_bt <- lmer(as.formula(getFormula(lmer_rlt, lmer=TRUE)), data = data_bt)  
        sd_res_bt <-  sigma(lmer_bt)  
        sd_RE_bt <-  lmer_bt@theta * sd_res_bt
        c(fixef(lmer_bt), sigma_RE = sd_RE_bt, sigma_res = sd_res_bt)  # get fixed coefficients 
    }) 
}
```

Inside `do(num_do)({...})`, observe the additional random component for the random effects, which is a vector of random intercepts of length = number of random effect variable.  We assign these random draws to individual units.  


### Apply bootstrapping to ChickWeight data  {-} 

####  Model 1 {-} 

- Model 1: linear time trend $$weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + \varepsilon_{ijt}$$ 

```{r}
ChickWeight2 <- ChickWeight

# model 1 
model_1 <- lm( weight ~ Diet*Time - Diet, data = ChickWeight2) 

rlt_model_1_bt <- run_ols_boot(model_1, num_do = 2000)

obs_model_1 <- tidy(model_1)  # summary of the original OLS estimates 

bt_sd_1 <- apply(rlt_model_1_bt, 2, sd) # calculate bootstrap standard errors 
bt_mode1_1 <- cbind(coeff = obs_model_1$estimate, 
                 sd = bt_sd_1, 
                 tstat = obs_model_1$estimate/bt_sd_1) 

# OLS estimate with statistical inferences by statistic theory
obs_model_1

# OLS estimate with statistical inferences by bootstrapping
bt_mode1_1 


ci_ver1(rlt_model_1_bt)
ci_ver2(obs_model_1$estimate, bt_sd_1, model_1$df.residual)
bt_p_val(obs_model_1$estimate, rlt_model_1_bt)$p_val 

# histogram
coeff_bt_histogram(obs_model_1$estimate, rlt_model_1_bt, centering=FALSE)
```

####  Model 2 {-} 

- Model 2: linear time trend with Chick fixed effects $$ weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + \alpha_i+ \varepsilon_{ijt}$$

```{r}
# model 2 
model_2 <- lm( weight ~ Diet*Time - Diet + Chick, data = ChickWeight2) 
rlt_model_2_bt <- run_ols_boot(model_2, num_do = 2000)

obs_model_2 <- tidy(model_2)  # summary of the original OLS estimates 

bt_sd_2 <- apply(rlt_model_2_bt, 2, sd) # calculate bootstrap standard errors 
bt_mode1_2 <- cbind(coeff = obs_model_2$estimate, 
                 sd = bt_sd_2, 
                 tstat = obs_model_2$estimate/bt_sd_2) 

# OLS estimate with statistical inferences by statistic theory
obs_model_2

# OLS estimate with statistical inferences by bootstrapping
bt_mode1_2

ci_ver1(rlt_model_2_bt)
ci_ver2(obs_model_2$estimate, bt_sd_2, model_2$df.residual)
bt_p_val(obs_model_2$estimate, rlt_model_2_bt)$p_val 

obs_model_2_time <- obs_model_2 %>% filter(grepl("Intercept", term) | grepl("Time", term)) 
obs_model_2_chick <- obs_model_2 %>% filter(grepl("Chick", term)) 
rlt_model_2_bt_time <- rlt_model_2_bt %>% dplyr::select(contains("Intercept"), contains("Time")) 
rlt_model_2_bt_chick <- rlt_model_2_bt %>% dplyr::select(contains("Chick")) 

coeff_bt_histogram(obs_model_2_time$estimate, rlt_model_2_bt_time, centering=FALSE)
coeff_bt_histogram(obs_model_2_chick$estimate[1:6], rlt_model_2_bt_chick[,1:6], centering=FALSE)
```

####  Model 3 {-} 

- Model 3: linear time trend with Chick random effects $$ weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + v_{ijt}, \:\: v_{ijt} = \alpha_i+ \varepsilon_{ijt}$$

```{r}
# model 3
model_3 <- lmer( weight ~ Diet*Time - Diet + (1 | Chick), data = ChickWeight2) 
rlt_model_3_bt <- run_lmer_boot(model_3, num_do = 2000)

obs_model_3 <- tidy(model_3)  # summary of the original lmer estimates 

bt_sd_3 <- apply(rlt_model_3_bt, 2, sd) # calculate bootstrap standard errors 
bt_mode1_3 <- cbind(coeff = obs_model_3$estimate, 
                 sd = bt_sd_3, 
                 tstat = obs_model_3$estimate/bt_sd_3) 
 
# random model estimate with statistical inferences by statistic theory (default)
obs_model_3

# random model estimate with statistical inferences by bootstrapping
bt_mode1_3

ci_ver1(rlt_model_3_bt)
ci_ver2(obs_model_3$estimate, bt_sd_3, (nrow(ChickWeight2) - nrow(obs_model_3)))
bt_p_val(obs_model_3$estimate, rlt_model_3_bt)$p_val 
coeff_bt_histogram(obs_model_3$estimate, rlt_model_3_bt, centering=FALSE)
```


<a href="../4-2-boot.html">Go back</a>
