
```{r, echo=FALSE }
suppressWarnings({
  suppressMessages({
    library(dplyr)
    library(knitr)
  })  
})
```

# Essentials {#essentials}

`r Sys.Date()`: <span style="color:red">*VERY Preliminary!*</span>

This section provides an overview of the essential concepts for manipulating data and programming in R. 


## Cheatsheets 

Cheatsheets are useful for glancing at various functions.    

* [Base R](http://github.com/rstudio/cheatsheets/raw/master/source/pdfs/base-r.pdf)

* [RStudio IDE](https://www.rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf)

* [dplyr](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf)

* [ggplot2](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)


## Data types

### Atomic 

In most cases, each atomic element has a type (mode) of;

* `numeric`: number

* `logical`: TRUE or FALSE (T or F for shortcuts)

* `character`: character string

* `factor`: a level of categorical variable   

Other types include [date](http://www.statmethods.net/input/dates.html) and nonexistent `NULL`. The factor is also a *class* of its own, meaning that many R functions apply operations that are specific to the *factor class*.  

```{r} 
# assess objects 123, "abc", and TRUE for their types 
str(123)  # str() returns the structure
str("abc")
str(TRUE)
c(is.numeric(123), is.numeric("abc"), is.numeric(TRUE))
c(is.logical(123), is.logical("abc"), is.logical(TRUE))
c(is.character(123), is.character("abc"), is.character(TRUE))

# "<-" means an assignment from right to left
factor1 <- as.factor(c(1,2,3)) # Looks like numeric but not
factor1
factor2 <- as.factor(c("a","b","c"))  # Looks like characters but not
factor2
factor3 <- as.factor(c(TRUE,FALSE,T)) # Looks like logicals but not
factor3
c(is.factor(factor1[1]), is.factor(factor2[1]), is.factor(factor3[1]))

# Extract the first element (factor1[1] etc.) 
factor1[1]
factor2[2]
factor3[3]
```

`NULL` has zero-length. Also, empty numeric, logical, and character objects have zero-length.  
```{r}
length(NULL) 
length(numeric(0))  # numeric(N) returns a vector of N zeros 
length(logical(0))  # logical(N) returns a vector of N FALSE objects
length(character(0)) # character(N) returns a vector of N "" objects
```

Each **vector** has a type of `numeric`, `logical`, `character`, or `factor`.
Each **matrix** has a type of `numeric`, `logical`, or `character`.
A **data frame** can contain mixed types across columns where each column (e.g., a variable) has a type of `numeric`, `logical`, `character` or `factor`. 

```{r}
vector1 <- c(1, NA, 2, 3) # read as numeric
vector1
vector2 <- c(TRUE, FALSE, T, F) # read as logical
vector2
vector3 <- c(1, NA, "abc", TRUE, "TRUE") # read as character
vector3
vector4 <- as.factor(c(1, NA, "abc", TRUE, "TRUE")) # read as factor 
vector4
```

```{r}
matrix1 <- matrix(c(1:6), nrow = 3) # read as numeric
matrix1
matrix2 <- matrix(c(TRUE,FALSE,rep(T,3),F), nrow = 3)  # read as logical
matrix2
matrix3 <- matrix(c(1,2,3,"a","b","abc"), nrow = 3) # read as character
matrix3

df1 <- data.frame(
        num  = c(1,2,3),           # read as numeric
        fac1 = c("a","b","abc"),   # read as factor
        logi = c(TRUE, FALSE, T),  # read as logical
        fac2  = c(1,"a",TRUE)      # read as factor
       )  
df1
df1$num   # "$" symbol is used to extract a column 
df1$fac1  # character type is converted into a factor 
df1$logi
df1$fac2  # mixed types within a column is converted into a factor

# additional argument "stringsAsFactors = FALSE" preserves character types.
df2 <- data.frame(
        num  = c(1,2,3),           # read as numeric
        char = c("a","b","abc"),   # read as character
        logi = c(TRUE, FALSE, T),  # read as logical
        fac2  = as.factor(c(1,"a",TRUE)),      # read as factor
        stringsAsFactors = FALSE
       )  
df2
df2$num  
df2$char
df2$logi
df2$fac2
```

### Factor

A factor object is defined with a set of categorical levels, which may be labeled. The levels are either  **ordered** (defined by `ordered()`) or **unordered** (defined by `factor()`). 
Ordered factor objects are treated in the specific order by certain statistical and graphical procedures.    

```{r}
# We will convert the columns of df into factors   
df <- data.frame(
      fac1 = c(0,1,1,4,4,2,2,3),
      fac2 = c(1,2,3,1,1,2,2,3),
      fac3 = c(4,2,3,4,4,2,2,3)
      )

# convert fac1 to ordered factors
df$fac1 <- ordered(df$fac1,
                  levels = c(0,4,3,2,1) # defines the order
                  ) 
df$fac1
summary(df$fac1)  # gives the table of counts for each level



# convert fac2 to unordered factors with labels
df$fac2 <- factor(df$fac2,
                  levels = c(1,2,3), #  no particular order
                  # attach labels to factors: 1=red, 2=blue, 3=green
                  labels = c("red", "blue", "green")
                  ) 
df$fac2
summary(df$fac2)                

# convert fac3 to ordered factors with labels
df$fac3 <- ordered(df$fac3,
                  levels = c(2,3,4),
                  # attach labels to factors: 2=Low, 3=Medium, 4=High
                  labels = c("Low", "Medium", "High")
                  ) 
df$fac3
summary(df$fac3)
```

### Matrix

`matrix()` defines a matrix from a vector. The default is to arrange the vector by column (`byrow = FALSE`).  

```{r}
# byrow = FALSE  (the default)
matrix(data = c(1:6), nrow = 2, ncol = 3, byrow = FALSE, dimnames = NULL)

# byrow = TRUE 
matrix(data = c(1:6), nrow = 2, ncol = 3, byrow = TRUE, dimnames = NULL)

# give row and column names to a matrix 
mat1 <- matrix(data = c(1:6), nrow = 2, ncol = 3, byrow = FALSE, 
       dimnames = list(c("r1","r2"), c("c1","c2","c3")))
mat1
dim(mat1)  # dimension: row by column
colnames(mat1)
rownames(mat1)

colnames(mat1) <- c("v1","v2","v3")  # change column names by assignment "<-" 
mat1

# R makes a guess when only nrow or ncol is supplied
matrix(data = c(1:6), nrow = 2)
matrix(data = c(1:6), ncol = 3)

# combine matrices by column via "cbind()" or by row via "rbind()"
cbind(mat1,mat1)
rbind(mat1,mat1)

```

There are recycling rules in R. 

```{r}
# the vector shorter than the length of all elements of a matrix
matrix(data = c(1:4), nrow = 2, ncol= 3)

# R treats a scaler as a vector of length that conforms cbind() or rbind()  
cbind(mat1, colA = 1)
rbind(mat1, rowA= 1, rowB= 2, rowC= 3)

```

To replace elements of a matrix, we can use assignment operator `<-`. 
```{r}
mat1[1,1] <- 10
mat1

mat1[,2] <- c(7,8)
mat1

mat1[,1] <- 0  # recycling rule
mat1
```

Matrix allows for easy extraction for rows and columns separated by comma. 
```{r}
mat1
mat1[1, ]  # row = 1 and all columns
mat1[, 1]  # all rows and col = 1 

mat1[c(TRUE,FALSE),]  # by a logical vector 
mat1[, c(TRUE,FALSE)]

mat1[2,3]  # row = 2 and col = 3
mat1[1:2, 2:3]  # row = 1:2 and col = 2:3
mat1[1:2, 2:3][2,2]  # subset of a subset
mat1[, 1][2]  # vector extraction is done with one-dimensional index  
```

Important: when a single row or column is extracted, it gets converted to a vector with no dimension. 
```{r}
mat1[1,]
is.matrix(mat1[1, ])  
dim(mat1[1,])
length(mat1[1, ])  

# to keep a row or column vector structure, use drop = FALSE
mat1[1,, drop = FALSE]
is.matrix(mat1[1,,drop = FALSE])  
dim(mat1[1,,drop = FALSE])
length(mat1[1,,drop = FALSE])  

mat1[,1, drop = FALSE]
is.matrix(mat1[,1,drop = FALSE])  
dim(mat1[,1,drop = FALSE])
length(mat1[,1,drop = FALSE])  
```

Another way of extraction from a matrix is to use row or column names.
```{r}
mat1[,'v1']
mat1[,c('v1','v3')]
mat1['r2',,drop= FALSE]
```

`apply()` applies a function for a specified margin (dimension index number) of the matrix. 

```{r}
mat1
apply(mat1,1,mean)  # dimension 1 (across rows)
apply(mat1,2,mean)  # dimension 2 (across columns)

# one can write a custom function inside apply(). (called annonymous function)    
# Its argument corresponds to the row or column vector passed by apply(). 
apply(mat1,2, function(x) sum(x)/length(x) )  # x is the internal vector name

ans1 <- apply(mat1,2, function(x) {   
                         avg = mean(x)
                         sd = sd(x)
                          # return the results as a list
                         list(Avg = avg, Sd = sd)
                      }
        )

unlist(ans1[[2]])  # results for the second column  
unlist(ans1[[3]])  # results for the third column  
```


Arrays are a generalization of matrices and can be more than 2 dimension. 
```{r}
array(c(1:18), c(2,3,3))  # dimension 2 by 2 by 3

array(c(1:9), c(2,3,3))  # R recycles the vector 
```

### Data Frame
A data frame is similar to a matrix, but it accepts multiple types (modes) of variables across columns (e.g., a dataset in typical data analysis programs like SAS, SPSS, Stata etc.). In some cases matrices and data frames may be treated interchangeably, but generally they need to be distinguished. Data manipulation functions are often written for data frames, while some base R functions are written for matrices.    

```{r}
mymat1 <- matrix(data = c(1:6), nrow = 2, ncol = 3, 
       dimnames = list(c("r1","r2"), c("c1","c2","c3"))) 
mymat1

class(mymat1)
colnames(mymat1)
names(mymat1)

mydf1 <- data.frame(
                mymat1, 
                num  = c(1,2),         
                fac1 = c("a","abc"),    
                logi = c(TRUE, FALSE), 
                fac2  = c(1,"a")   
              )  
mydf1

class(mydf1)
colnames(mydf1)
names(mydf1)   # colnames and names are the same 
```

Extracting elements from a data frame is similar to extracting from a matrix, but there are a few additional methods.
```{r}
mydf1[1,]   # row = 1 and all columns 
mydf1[,1]   # all rows and col = 1

# data frame preserves dimension while extracting a row but not a column
dim(mydf1[1,])  
dim(mydf1[,1])  
dim(mydf1[,1,drop=FALSE])   # use drop = FALSE to keep a column vector
mydf1[,1,drop=FALSE]

mydf1[, c('c1','num','logi')]
class(mydf1[, c('c1','num','logi')])

# extraction by column name with "$" symbol:  df$varname  
mydf1$c1
dim(mydf1$c1)
# one can use quote ' ' or " " as well 
mydf1$'c1'

# similarly, extraction by column name with [[ ]]: df[['varname']]    
mydf1[['c1']]
dim(mydf1[['c1']])

# or by index 
mydf1[[1]]


# [[ ]] method is useful when passing a variable name as a string
set_to_na <- function(df, var) {
                df[[var]] <- NA  
                df 
              }
mydf1
mydf2 <- set_to_na(mydf1, "c2") 
mydf2

# add a variable 
mydf1$newvar <- c(4, 4)
mydf1

mydf1$newvar2 <- mydf1$c2 + mydf1$c3
mydf1
```

`apply()` may not work well with data frames since data frames are not exactly matrices. We can use simplified apply `sapply()` or list apply `lapply()` instead.  

```{r}
mydf1 

# sapply() 
idx_num <- sapply(mydf1, is.numeric) 
idx_num
apply(mydf1[,idx_num], 2, mean)
sapply(mydf1[,idx_num], mean)

# lapply() 
idx_num2 <- unlist(lapply(mydf1, is.numeric)) 
idx_num2
unlist(lapply(mydf1[,idx_num2], mean))
```

### List

A list is an ordered collection of (possibly unrelated) objects.  The objects in a list are referenced by [[1]], [[2]], ..., or [['var1']], [['var2']], ... etc.   

```{r}
mylist1 <- list(v1 = c(1,2,3),
             v2 = c("a","b"),
             v3 = factor(c("blue","red","orange","yellow")),
             v4 = data.frame( u1 = c(1:3), u2 = c("p","q","r"))
             )
mylist1 

# extraction
mylist1[[1]]             
mylist1[["v2"]]             
mylist1$v3      
mylist1$v4$u2

# assignment
mylist1$v5 <- c("a",NA)
mylist1$v5 

# a list can be nested 
mylist1$v6 <- list(y1 = c(2,9), y2 = c(0,0,0,1))
mylist1$v6
```

`lapply()` is very versatile since the items in a list can be completely unrelated. 

```{r}
unlist(lapply(mylist1, class))
unlist(lapply(mylist1, attributes))  # some variables have attributes

lapply(mylist1, function(x) {
                  if (is.numeric(x)) return(summary(x))
                  if (is.character(x)) return(x)
                  if (is.factor(x)) return(table(x))
                  if (is.data.frame(x)) return(head(x))
                  if (is.list(x)) return(unlist(lapply(x,class)))
                }
)
```

## Programming

### Operator

```{r, echo=FALSE}
tbl_operator <- data.frame(
                  Operation = c("x + y", "x - y", "x * y", "x / y", "x ^ y",
                                "x %% y", "x %/% y", "x == y", "x <= y", "x >= y",
                                "x && y", "x || y", "x & y", "x | y", "!x"),
                  Description = c("Addition","Subtraction","Multiplication","Division","Exponentiation",
                                  "Modular arithmatic","Integer division","Test for equality",
                                  "Test for less than or equal to","Test for greater than or equal to",
                                  "Boolean AND for scalars", "Boolean OR for scalers",
                                  "Boolean AND for vectors", "Boolean OR for vectors",
                                  "Boolean negation")
                )
```

`r tbl_operator %>% kable(caption = "Basic R Operators")`

source: [@Matloff2011]


### If else

```{r}
if (1 > 0) {
  print("result: if")
} else {
  print("result: else")
}

# {} brackets can be used to combine multiple expressions 
# They can be skipped for a single-expression if-else statement. 
if (1 > 2) print("result: if")  else print("result: else")

ifelse(c(1,2,3) > 2, 1, -1)  # return 1 if TRUE and -1 if else  

Sys.time()
time <- Sys.time()
hour <- as.integer(substr(time, 12,13))
# sequential if-else statements
if (hour > 8 & hour < 12) {
    print("morning")
} else if (hour < 18) {
    print("afternoon")
} else {
    print("private time")
}
```

### Loop

```{r}
for (i in 1:3) {
  print(i)
}

for (i in c(1,3,5)) print(i)

i <- 1 
while (i < 5) {
  print(i) 
  i <- i + 1
}
```


### Function

We can avoid repeating ourselves with writing similar lines of codes if we turn them into a function. Functions contain a series of tasks that can be applied to varying objects such as different vectors, matrices, characters, data frames, lists, and functions.

A function consists of input arguments, tasks (R expressions), and output as an object (e.g. a vector, matrix, character, data frame, list, or function etc.). It can be named or remain anonymous (typically used inside a function like `lapply()`). 

```{r}
##  name_to_be_assigned  <- function(input args) {
##                            tasks 
##                           }
```

The output of the function, aside from those that are printed, saved, or exported, is the very last task (expression). If variable `result` is created inside the function, having `result` at the very end will return this item as an output. When multiple objects are created, it is often convenient to return those as a *list*. Use `return()` to return a specific item in the middle of the function and skip the rest of the evaluations.  For checking errors and halting evaluations, use `stop()` or `stopifnot()`.           



```{r}
myfun1 <- function() print("Hello world")  # just returning "Hello world"
myfun1()  

myfun2 <- function(var) var^2 
myfun2(var = 3)

myfun3 <- function(var = 1)  ifelse(var>0, log(var), var) 
myfun3()   # default argument is var = 1
myfun3(2)

myfun4 <- function(x1, x2) {
            if (!is.numeric(x1) | !is.numeric(x2)) stop('demo of error: numeric args needed')
            x1*x2
          }
# try(myfun4(1, "a"))  
myfun4(4, 3)
```


### Environment

A function, formally known as a *closure*, consists of its arguments (called *formals*), a body, and an *environment*. An *environment*  is a collection of existing objects at the time when the function is created. Functions created at the *top level* have `.GlobalEnv` as their environments (R may refer to it as `R_GlobalEnv` as well).  


```{r}
environment()  # .GlobalEnv  (or R_GlobalEnv) is the top-level environment 
f1 <- function(arg1) environment()  
formals(f1)  # arguments of f1()
body(f1)     # body of f1()
environment(f1)  # environment of f1(), which is .GlobalEnv 
f1()  # inside f1 has its own enviornment 
```

A function can access to the objects in its environment (i.e., *global* to the function) and those defined inside (i.e., *local* to the function) and generally cannot overwrite the global objects. It allows for using common names such as "x1", "var1" etc. defined inside functions, but those objects are only accessible within the function. 

```{r}
a <- NULL  # object named "a" in .GlobalEnv
f2 <- function() {
        a <- 1  # object named "a" in an environment inside f2
        print(a)
        environment()
      }
f2()  #  one instance creating an environment   
f2()  #  another instance creating another environment
a  # stays NULL 

ls()  # ls() shows all objects of an environment (here .GlobablEnv)
rm(list = ls())  # rm() removes items of an  environment (here .GlobablEnv)
ls()  # all gone in GlobalEnv
```

Using  global assignment `<<-` operator,  one can bend this general rule of not affecting global objects. This can be useful when it is desirable to make certain objects accessible across multiple functions without explicitly passing them through arguments.  

```{r}
a <- NULL
b <- NULL
f1 <- function() {
        a <<- 1  # global assignment
        # another way to assign to GlobalEnv 
        assign("b", 2, envir = .GlobalEnv)  
      }
f1() 
a
b

a <- 2
f2 <- function() {
        # Since there is no "a" local to f2, R looks for "a" 
        # in a parent environment, or .GlobalEnv  
        print(a) 
        
        # g() assigns a number to "a" in g()'s environment 
        g <- function() a <<- 5
        
        a <- 0  #  object called "a" local to f2   
        print(a)
        
        # g() updates only the local "a" to f2(), but not "a" in GlobalEnv 
        # R's scope hierarchy starts from local to its environment  
        g()   
        print(a)
      }
a <- 3

# the first "a" is in .GlobalEnv when f2() is called
# the second "a" is local to an instace of f2()
# the third "a" is the updated version of the local "a" by g() 
f2() 
a   # object "a" in GlobalEnv: unchanged by g()
```

It is convenient to use `<<-` if you are sure about which object to overwrite. Otherwise, the use of `<<-` should be avoided.    

### Debugging 

`browser()` and `traceback()` are common debugging tools. A debugging session starts where `browser()` is inserted and allows for a line by line execution onward. Putting `browser()` inside a loop or function is useful because it allows for accessing the objects at a particular moment of execution in its environment.  After an error, executing `traceback()` shows  at which process the error occurred. Other tools include `debug()`, `debugger()`, and `stopifnot()`. 


### Stat func.

```{r, echo=FALSE}
tbl_statfunc <- data.frame(
                  Distribution = c("Normal", "Chi square", "Binomial"),
                  Density_pmf =c("dnorm( )","dchisq( )", "dbinom( )"),
                  cdf = c("pnorm( )", "pchisq( )", "pbinom( )"),
                  Quantiles = c("qnorm( )","qchisq( )", "qbinom( )"),
                  Random_draw = c("rnorm( )", "rchisq( )","rbinom()")
                )
```

`r tbl_statfunc %>% kable(caption = "Common R Statistical Distribution Functions")`

source: [@Matloff2011]

### String func.

R has built-in string manipulation functions. They are commonly used for;

* detecting a certain pattern in a vector (`grep()` returning a location index vector, `grepl()` returning a logical vector)

* replacing a certain pattern with another (`gsub()`)

* counting the length of a string (`nchar()`) 

* concatenating characters and numbers as a string (`paste()`, `paste0()`, `sprintf()`)

* extracting a segment of a string by character position range (`substr()`)

* splitting a string with a particular pattern (`strsplit()`)

* finding a character position of a pattern in a string (`regexpr()`)

```{r}
oasis <- c("Liam Gallagher", "Noel Gallagher", "Paul Arthurs", "Paul McGuigan", "Tony McCarroll") 
grep(pattern = "Paul", oasis)

grepl(pattern = "Gall", oasis)

gsub("Gallagher", "Gallag.", oasis)

nchar(oasis)

paste(oasis)

paste(oasis, collapse=", ")

sprintf("%s from %d to %d", "Oasis", 1991, 2009)

substr(oasis, 1, 6)

strsplit(oasis, split=" ")  # split by a blank space

regexpr("ll", oasis[1])[1]
```

Common regular expressions used in R include;

* `"[char]"` (any string containing either "c", "h", "a", or "r" )

* `"a.c"` (any string containing "a" followed by any letter followed by "c")

* `"\\."` (any string containing symbol ".").   

```{r}
grepl("[is]", oasis)

grepl("P..l", oasis)

grepl("\\.", c("Liam", "Noel", "Paul A.", "Paul M.", "Tony"))
```


### Set func.
The functions for common set operations include `union()`, `intersect()`, `setdiff()`, and `setequal()`. The most commonly used function is `%in%` operator; `X %in% Y` returns a logical vector indicating whether an each element of `X` is a member of `Y`.        
```r
c(1,2,3,4,5) %in% c(3,2,5)
c("a","b","t","s") %in% c("t","a","a")
```


## Housekeeping

### Working directory

`getwd()` returns the current working directly.  `setwd(new_directory)`  sets a specified working directory. 

### R session

`sessionInfo()` shows the current session information. In RStudio, `.rs.restartR()` restarts a session. 

### Save & load 

R objects can be saved and loaded by `save(object1, object2, ..., file="file_name.RData")` and `load(file="file_name.RData")`. A `ggplot` object can be save by `ggsave("file_name.png")`.   


### Input & Output

A common method to read and write data files is `read.csv("file_name.csv")` and `write.csv(data_frame, file = "file_name.csv")`. `scan()` is a more general function to read data files and interact with user keyboard inputs. `file()`  is also a general functions to read data through *connections*, which refer to R's mechanism for various I/O operations. `dir()` returns the file names in your working directory.     

A useful function is `cat()`, which can print a cleaner output to the screen, compared to `print()`.
```{r}
print("example")

cat("example\n")  # end with \n

cat("some string", c(1:4), "more string\n")
cat("some string", c(1:4), "more string\n", sep="_")
```


### Updating

R needs regular updates for R distribution, individual R packages, and RStudio. Generally, updating once or twice a year would suffice.  For updating RStudio, go to `Help` and then `Check for Updates`. Also, RStudio also makes it easy to update packages; go to `Tools` and the `Check for Package Updates`.  Do these updates when you have time or you know that you need to update a particular package; updating R and R packages can be trickier than it seems.    

```{r}
# check R version
getRversion()
version

# check installed packages
## installed.packages()

# list all packages where an update is available
## old.packages()

# update all available packages of installed packages 
## update.packages()

# update, without prompt
## update.packages(ask = FALSE)
```

For windows users, one can automate the process using `installr` package. 
```{r}
## --- execute the following ---
## install.packages("installr") # install
## setInternet2(TRUE) # only for R versions older than 3.3.0
## installr::updateR() # updating R.
```

<!-- not sure if this works
Another way to automate is discussed [here](http://bioinfo.umassmed.edu/bootstrappers/bootstrappers-courses/courses/rCourse/Additional_Resources/Updating_R.html).
```{r}
### get packages installed
## packs = as.data.frame(installed.packages(.libPaths()[1]), stringsAsFactors = F)

### and now re-install install packages using install.packages()
## install.packages(packs$Package)
```
--> 

Sometimes, you can accidentally corrupt sample datasets that come with packages. To restore the original datasets, you have to remove the package by `remove.packages()` and then install it again. Use `class()`, `attributes()`, and `str()` to check for any unrecognized attributes attached to the dataset. Also, if you suspect that you have accidentally corrupted R itself, you should re-install the R distribution.  




