---
title: "Demo. Mixed Effects Models and LSMEANS"
output: html_document

---

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(lsmeans)

# make a copy of dataset
ChickWeight2 <- ChickWeight
```

Mixed effects models employing fixed and random parameters are popular in the analysis of experimental data.  In some disciplines, there is a strong tradition of summarizing results with "LSMEANS", a [SAS's terminology](http://andrey-myslivets.tumblr.com/post/92630498431/en).  

Here we will briefly  demonstrate mixed effect estimation and
[lsmeans](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=0ahUKEwjk4-XQo_LbAhVk4oMKHXY4ACUQFggoMAA&url=https%3A%2F%2Fcran.r-project.org%2Fweb%2Fpackages%2Flsmeans%2Fvignettes%2Fusing-lsmeans.pdf&usg=AOvVaw2kxQsw-sISiXmxlP2kTzj4) in R.   

# Mixed Effect Models {-}

Let $weight_{ijt}$ be the weight of chick $i$ in Diet group $j$ observed in time $t$.  Consider modeling the effect of Diet on weight. (If you are new to R estimation syntax, [reading this](http://conjugateprior.org/2013/01/formulae-in-r-anova/)
 may be helpful.) 

### Model 1 linear time trend $\beta_{1}$, Diet fixed effects $\alpha_{j}$,  Chick random effects $u_i$ {-}

$$ weight_{ijt} = \alpha_{j} + \beta_{1}\: time_t + v_{ijt}, \:\: v_{ijt} = u_i+ \varepsilon_{ijt}$$
where $\varepsilon_{ijt}$ is an error term. 

```{r}
model_1 <- ChickWeight2 %>% 
  with(lmer(weight ~  Diet + Time  + (1 | Chick))) 

# marginal effects 
summary(model_1)

# lsmeans: projected mean for each Diet 
lsmeans::lsmeans(model_1, specs=c("Diet"))
```

### Model 2: discrete time fixed effects $\beta_{t}$, Diet fixed effects $\alpha_{j}$ and Chick random effects $u_i$ {-}

$$ weight_{ijt} = \alpha_{j} + \beta_{t} + v_{ijt}, \:\: v_{ijt} = u_i+ \varepsilon_{ijt}$$

```{r}
# add a factor time variable
ChickWeight2 <- ChickWeight2 %>% mutate(Time_fac = as.factor(Time)) 

model_2 <- ChickWeight2 %>%
  with(lmer(weight ~  Diet + Time_fac + (1 | Chick)))

# marginal effects 
summary(model_2)

# lsmeans: projected mean for each Diet 
lsmeans::lsmeans(model_2, specs=c("Diet"))

# lsmeans: projected mean for each combination of Time_fac and Diet 
lsmeans::lsmeans(model_2, specs=c("Time_fac","Diet"))
```


### Model 3: Diet fixed effects on linear-time growth rates $\beta_{1j}$,  Chick random effects $u_i$ {-}

$$ weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + v_{ijt}, \:\: v_{ijt} = u_i+ \varepsilon_{ijt}$$

Note: common intercept at Time=0 across Diet types.

```{r}
# make a copy of dataset
ChickWeight2 <- ChickWeight

model_3 <- ChickWeight2 %>% 
  with(lmer(weight ~ Diet*Time - Diet + (1 | Chick))) 

# marginal effects 
summary(model_3)

# lsmeans: projected mean for each Diet at given Time values 
lsmeans::lsmeans(model_3, specs=c("Diet"), at=list(Time=c(5, 10, 20)))
```

### Model 4: Diet fixed effects on quadratic-time growth rates $\beta_{1j}$,  Chick random effects $u_i$ {-}

$$ weight_{ijt} = \alpha_{0} + \beta_{1j}\: time_t + \beta_{2j}\: time_t^2 + v_{ijt}, \:\: v_{ijt} = u_i+ \varepsilon_{ijt}$$

Note: common intercept at Time=0 across Diet types.

```{r}
model_4 <- ChickWeight2 %>%
  with(lmer(weight ~ Diet:Time + Diet:I(Time^2) - Diet  + (1 | Chick)))

# marginal effects
summary(model_4)

# lsmeans: projected mean for each Diet at given Time values 
lsmeans::lsmeans(model_4, specs=c("Diet"), at=list(Time=c(5, 10, 20)))
```


# Visualizing Model Fit {-}

### Raw data {-}
```{r}
# raw data plot 
base_plot <- 
  ChickWeight2 %>%
    ggplot(aes( x = Time, 
                y = weight, 
                color = Diet)) + 
    geom_jitter(size =.75, alpha=.5) 

base_plot
```

### Model 1 fit (approx.) {-}

```{r}
# overlay model_1, approximated by  y ~ x
base_plot +
  geom_smooth(
    aes(x = Time, 
        y = fitted.values(model_1),
        color = Diet),
    method='lm', formula = y ~ x, se=FALSE)
```

### Model 2 fit (approx.) {-}
```{r}
# overlay model_2, approximated by loess 
base_plot +
  geom_smooth(
    aes(x = Time, 
        y = fitted.values(model_2),
        color = Diet), se=FALSE)
```

### Model 3 fit (approx.) {-}
```{r}
# overlay model_3, approximated by  y ~ x
base_plot +
  geom_smooth(
    aes(x = Time, 
        y = fitted.values(model_3),
        color = Diet),
    method='lm', formula = y ~ x, se=FALSE)
```

### Model 4 fit (approx.) {-}
```{r}
# overlay model_4, approximated by  y ~ x + x^2 
base_plot +
  geom_smooth(
    aes(x = Time, 
        y = fitted.values(model_4),
        color = Diet),
    method='lm', formula = y ~ x + I(x^2), se=FALSE)
```


